Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-29 tagless/tagless-final.tex
make many edits to sections 1,2,3.  section 4 to come later.

@@

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of implicit representations of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\oleg{say something about type functions, GADTs and the way to get
  around them?}\jacques{not in the abstract?}
\end{abstract}

@@

See Section 1 of Walid02. Our subject area is that of a definitional DSL
interpreter.
Define Staging, meta-language and object language. (Walid uses
`subject' language).

@@

The first attempt at the interpreter, the function eval. It takes the
object language term such as |test1| and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ -> List.hd env
  | V (VS v) -> eval0 (List.tl env) (V v)
  | B b -> b 
  | L e -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

Suppose our meta-language (the language in which we wrote test1
and that interpreter) were untyped. The above code would be acceptable.
The |L e| line exhibits interpretative overhead (see descr in Walid02
paper): |eval0 env' e| will be executed every time the result of
evaluating |L e| is applied. Staging can be used to remove the
interpretative overhead. Cite some paper -- see Walid02, Sec 1.1 and
1.2. They say it well.

If we use OCaml or some other typed language as the meta-language, 
the function |eval0| is ill-typed. The line |B b|
tells that the function returns a boolean, whereas the next line says
the result of |eval0| is a function. Clearly, the result of the
function can't be both of these types. 
A related problem is the type of the env: since variables
in the language may be either of boolean or function types, a regular
OCaml list cannot be used for holding the corresponding values.

The solution is to introduce a Universal type (see Walid02 Section
1.3) which contains both of these:

@@

We can then write the interpreter, which is accepted by the typecheker
\begin{code}
  let rec eval env = function
  | V VZ -> List.hd env
  | V (VS v) -> eval (List.tl env) (V v)
  | B b -> UB b
  | L e -> UA (fun x -> eval (x::env) e)
  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code}
and whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term:

@@

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, |List.hd| and
|List.tl| above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is, the first
operand is not a function.

Suppose our object language were typed, too. Then terms like
|A (B true, B false)| would be ill-typed. Thus, if our evaluator
accepted only typed terms, then the missing alternatives in the |A|
line would not occur. Alas, what is obvious to us is not obvious to
the meta-language. There is still a pattern match, and we still have
to attach and eliminate the UB and UA tags, even when there is no
longer a need for them. See Walid02, Sec 1.4 for more detail.

Of course there is also the issue how to design a type system to type
terms like |test1| (GADT seem necessary). Also, how to design the
typesystem to keep track of open/closed status of terms, so we can
statically assure that only closed terms are passed to eval and that
the empty list checking in |List.hd| and |List.tl| can be eliminated
as well. The latter is even more interesting problem requiring even
more advanced type systems (cite env classifiers, Nanevski, Rowan
Davies, Pitts).
(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
\jacques{I hope that is the right Davies paper}

The problem is also outlined in Simon Peyton-Jones GADT paper, but
only in the context of a first-order language

In short: \oleg{ (put into really introduction or the conclusions?) }
It is not satisfactory
to use the Universal type because tagging hurts performance and
necessitates partial pattern matching in the metalanguage, even for a
well-typed object term.

% \subsection{Other solutions}
% The tag problem motivated GADTs (SPJ, Xi) and dependent types
% (WalidICFP02, Concoqtion). Or, perhaps just refer to related work?
% But we need to say something here, to give the reader the feeling of
% how advanced solutions were proposed and how rarely they are implemented.

See the Related Work section (\ref{related}) for a discussion of the other
solutions to this problem.
\oleg{I feel we need to give some details of the other solutions here, 
to give the reader the feeling of
how advanced solutions were proposed and how rarely they are
implemented.}
\jacques{Sure, we can refer to this section from the Related Work section
for details.}

\subsection{The `Final' Proposal}\label{ourapproach}

@@

\oleg{see tagless\_interp1.ml, module Tagless for the complete code.
If you change the code in here, please adjust the .ml file
accordingly. Let the paper and the accompanying code be in sync.}

\jacques{do we want to start using a lot of little files and
use includes to insure the code stays in synch?}
\oleg{Because the fragements are interrelated, we can't. We have to
  use sort of a `partial include', and I don't think that is easy. I
  have checked all the fragments so far.}

Now, let us define our object language (embedded DSL) in a different
way. Instead of using algebraic datatypes to encode the terms of the
language in the meta-language (reminiscent to the initial approach),
we will be using functions as term constructors (the final approach):

@@

That is all of the interpreter. Our sample term now reads
\begin{code}
  let testf1 = app (lam varZ) (b true)
\end{code}
which is almost the same as before, only written with lower-case
identifiers. To evaluate the term, we merely need to apply 
the term to the empty environment

@@

\oleg{
We should emphasize that not only our code can't have pattern-match
errors, that fact is apparent to the compiler itself. It is one thing
to write a code with pattern-matching and claim (and prove) that
missing cases cannot occur. However true that may be, the fact the
missing cases cannot occur is not evident to the compiler, which still
has to generate code to report missing cases or compile in the error
generation. That's where efficiency is lost, and Walid's papers on tag
elimination have numbers showing that performance is lost indeed. 
In our code, there is simply no pattern-mismatch opportunity: the
proof of no bad match is expressed in the code itself. }

In the above code, the interpreter is wired in, directly built into the
functions |b|, |app|, etc.  In the remainder of this paper, 
we explain how to abstract the interpreter,
to make it possible to `evaluate' the same term in many different
ways: interpretation, compilation, compute the size (or depth), etc.

@@

        We have realized that if we change the encoding of the term --
from data constructors to functions, we can encode a tagless
interpreter of a typed language in a typed language with no GADTs and
a very simple type system. The interpreter uses no universal type, no
pattern matching and it is patently obvious to the compiler that the
only non-termination can come from interpreting |fix|. If the source
language is strongly normalizing, our interpreter is total.

@@

We have implemented the above in (Meta)OCaml and Haskell.

What are our contributions:
\begin{enumerate}
\item we interpret/compile/partially-evaluate a typed source language
   from within a typed language.  The requirements of the host type
   system are \jacques{not so onerous?  simple?},
\item a simple, clean implementation,
\item a nicely comparable Haskell/MetaOCaml version,
\item a functorial Symantics that deals with uniformly across all
evaluators, from interpretation to partial evaluation,
\item clear and easy extensibility to more features,
\item only closed terms can be evaluated.
\item an approach to self-interpretation which is compatible with the
  above.
\end{enumerate}

Making sure that we interpret terms that are well-typed and closed --
and we can do that in far simpler type system (e.g., Haskell or
OCaml). It was thought that doing either requires quite advanced
extensions (GADT, Nanevski, Concoqtion).
\jacques{can we state fairly clearly what type system we need?  I mean,
Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
With our let-polymorphism trick for self-interpretation, what do we really
need?}

@@

Our DSL language to be embedded is simply-typed lambda-calculus
with fixpoint, integers, booleans and comparison, whose BNF grammar is

@@

This time we use higher-order abstract syntax (cite: but not
Elliott. Cite Dale Miller!) rather than de Bruijn indices.  The
language is close to the language of Xi03, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Gibonacci, factorial, and power.

@@

|test1| now reads |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For example, we can encode the classical |power| function, as well
as the partial application |\x -> power x 7| as

\begin{code}
testpowfix () = 
  lam (\x -> fix (\self -> lam (\n ->
        if_ (leq n (int 0)) (int 1)
            (mul x (app self (add n (int (-1))))))))
testpowfix7 () = 
  lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code}
The dummy unit |()| argument above is to avoid the monomorphism
restriction and so that the type of |testpowfix| and |testpowfix7|
will remain polymorphic in |repr|. Instead of supplying the dummy
argument, we could have given the terms explicit polymorphic
signature.  We however prefer if the typechecker of the meta-language
(Haskell) infers the types of the object terms for us. We could also
avoid the |()| argument by switching off the monomophism restriction
with a compiler flag.

Note that the subject language (DSL) is typed. If we make a mistake
(for example, replace |int 7| with |bool True| in |testpowfix7|, the
compiler will highlight that term and complain that it couldn't match
the expected |Int| against the inferred |Bool|). A type error
also results when entering |lam (\x -> app x x)|, with the compiler
complaining about an infinite type. The type-checker of the
meta-language will flag meaningless DSL (subject level) terms --
before any of their evaluation. If we forget, for example, |app| in
some of the terms above, we get an error message with a precise
location. So the type-checker also flags syntactically invalid
subject-level terms.\oleg{Should we perhaps move this up, in the
  intro? To justify why typed-metalanguages and the typed DSL are
  good. Or keep the explanation here but refer to it from the Introduction.}

We also embedded our DSL in (Meta)OCaml, replacing type classes with
functors, via a module type.  For this section, the following simple
signature will be sufficient:

@@

Some differences are in the additional type parameter |c| (the
environment classifier, Walid03) to be used later when we get to code
generation (required by MetaOCaml, part of the type of the
code). MetaOCaml is a call-by-value language, and thus we have to 
$\eta$-expand
the fix-point combinator and encapsulate the branches of the |if_|
into thunks to prevent their evaluation.

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
     if_ (leq n (int 0)) (fun () -> int 1)
         (fun () -> mul x (app self 
                               (add n (int (-1))))))))
 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
The dummy unit argument in |testpowfix| is an artefact of MetaOCaml
(which we will be using to compile our tests in Sections below),
again related to monorphism. To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means, the type
must be polymorhic. The value restriction then dictates 
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

This interpreter is patently tagless: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. We can run our
examples by instatiating the |EX| functor with |R|.
\oleg{Jacques: you mentioned "Boxes go Bananas" by Washburn and
  Weirich. Do you want to comment more about that paper here?}

\noindent
\begin{code}
  module EXR = EX(R);;
\end{code}

\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define a\\
|newtype R a = R a| and |instance Symantics R|
with the straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during compilation) and pattern-matching
against |R| cannot ever fail and so is operationally an identity and
compiled that way.

Also, note that the interpreter uses no Universal type nor pattern
matching, and it is obvious to the compiler that non-termination can
only result from interpreting |fix|.

@@

\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int) = 1
  let mul e1 e2 = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1

  let lam f = (f 0) + 1
  let app e1 e2 = e1 + e2 + 1
  let fix f = (f 0) + 1
end;;
\end{code}

@@

We can also write a ``compiler'' from our lambda calculus directly
into OCaml code, using MetaOCaml's staging facilities. Namely,
future-stage expressions of the type |t| are represented at the
present stage as values of the type |('a,t) code| where |'a| is the
environment classifier (Walid03, ESOP04). Code values are created by
a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
evaluated at the future stage. The \emph{escape} |.~e|, which must occur
within a bracket, specifies that the expression |e| must be evaluated
at the present stage; its result, which must be a code value, is
spliced in into the code being built. The \emph{run} form |.! e|
evaluates the code value, which involves run-time compilation and
linking of the future-stage code. The bracket, escape and run are akin to
quasi-quotation, unquotation and |eval| of Lisp.

% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code
  let int (x:int) = .<x>.
  let bool (b:bool) = .<b>.
  let add e1 e2 = .<.~e1 + .~e2>.
  let mul e1 e2 = .<.~e1 * .~e2>.
  let leq x y = .< .~x <= .~y >.
  let if_ eb et ee = 
    .<if .~eb then .~(et () ) else .~(ee () )>.

  let lam f = .<fun x -> .~(f .<x>.)>.
  let app e1 e2 = .<.~e1 .~e2>.
  let fix f = 
     .<let rec self n = .~(f .<self>.) n in self>.
end;;
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler attempts no optimizations at all, it simply produces
residual code. For example, the evaluation of our |test1|
\begin{code}
  let module E = EX(C) in E.test1 ();;
\end{code}
gives the value of the inferred type |('a, bool) C.repr| which looks
like |.<((fun x_6 -> x_6) (true))>.|. MetaOCaml can print the code
values, that is, the corresponding future-stage code.

\begin{comment}
\oleg{one may wish to give the folloiwng example too; the code
  expression can be somewhat abbreviated...}
\begin{code}
  let module E = EX(C) in E.testpowfix7;;
# - : ('_a, int -> int) C.repr =
.<fun x_1 ->
   (((fun x_2 ->
       let rec self_3 =
        fun n_4 ->
         ((fun x_5 -> if (x_5 <= 0) then 1 else (x_2 * (self_3 (x_5 + (-1)))))
           n_4) in
       self_3) x_1) 7)>.
\end{code}
\end{comment}

\noindent It is readily obvious that this compiler does not suffer
any interpretive overhead whatsoever. Indeed, the
  code produced for a function is simply |fun x_6 -> x_6| and has no 
  embedded calls to the evaluator: cf. the evaluator of DSL functions
  in Section~\ref{tagproblem}, the L-line.
The resulting code obviously contains no tags and no pattern-matching.

We have also implemented the ``compiler'' in Haskell. Since Haskell
has no (convenient, and typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with the
constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
meaning. Unlike our staging language (which is based on higher-order
abstract syntax), our bytecode, to be realistic, uses integer-named
variables. We then define 
\begin{code}
  newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for the creation of fresh variable
names. We define the compiler by making |C| to be the instance of the
class |Symantics|. The implementation is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying complete code gives more details.

\oleg{I wanted to avoid saying anything about GADTs here. There
  doesn't seem to be much point in getting into too long
  explanations...}

@@

Surprisingly, it is also possible to write a partial evaluator!  The
partial evaluator does have tags: whether a value is static or dynamic.
The pattern-matching on this \emph{phase} tag is always exhaustive; we still
have no universal type and no tags for object types.  The partial
evaluator uses the previously mentioned compiler to ``interpret'' dynamic
terms, and it uses the interpreter to reduce static terms.  In this sense, 
the partial evaluator is a modular extension to both the compiler and
the interpreter.

\subsection{Non-functorialness of functions problem}
\label{S:PE-problem}
We want to combine the interpreter |R| and the compiler |C| to produce
an (online) partial evaluator. \oleg{discuss if our PE online or
  offline: here or below. In the latter case refer to the discussion
  from here.} The partial evaluator (PE) is also an interpreter: it
evaluates a term either to immediate (or, `static' or present-stage)
value or a `dynamic', future-stage value, or `code' (of the sort
produced by the compiler C). The interpreter attempts to perform all
possible computations (i.e., computations whose operarands are
present-stage values) as the present-stage. Our first attempt is
straightforward: we introduce
\begin{code}
data P1 t = S1 (R t) (C t) | E1 (C t)
\end{code}
as a `direct sum' of the interpreter |R| and the compiler |C|. The
datatype |P1| has two \emph{phase} (not type!) tags |S1| 
(for static) and |E1| for dynamic. The static value has both the
present-stage and future-stage alternatives. The dynamic value does
not have the present-stage component. We define
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (S1 _ dyn) = dyn
abstr1 (E1 dyn)   = dyn
\end{code}
to extract the future-stage component. We then attempt to define the
interpreter |P1| compositionally -- and it works for the first-order
fragment of our DSL language.

@@

@@

@@

\end{document}

>>>>

\usepackage{natbib1}
\bibpunct();A{},
\let\cite=\citep

@@

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\oleg{say something about type functions, GADTs and the way to get
  around them?}\jacques{not in the abstract?}
\end{abstract}

@@

See Section 1 of Walid02. Our subject area is that of a definitional DSL
interpreter. 
Define Staging, meta-language and object language. (Walid uses
`subject' language).
\jacques{\emph{subject} and \emph{object} are both used in the current
write-up; should we define them to be equivalent, or just stick to one?}

@@

Let us attempt to implment an the interpreter, the function |eval0|. It takes
an object language term such as |test1| and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ      -> List.hd env
  | V (VS v)  -> eval0 (List.tl env) (V v)
  | B b       -> b 
  | L e       -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

Suppose our meta-language (the language in which we wrote test1
and that interpreter) were untyped. The above code would be acceptable.
The |L e| line exhibits interpretive overhead \cite{WalidICFP02}: 
|eval0 env' e| will be executed every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead (sections 1.1 and 1.2 of \cite{WalidICFP02} explain
this well).

If we use OCaml or some other typed language as the meta-language, 
the function |eval0| is ill-typed. The line |B b|
says that the function returns a boolean, whereas the next line says
the result is a function. Clearly, the result of |eval0|
can't be both of these types. 
A related problem is the type of the environment |env|: since variables
in the language may be either of boolean or function types, a regular
OCaml list cannot be used for holding the corresponding values.

The usual solution is to introduce a Universal type (section 1.3 of
\cite{WalidICFP02}) which contains both of these:

@@

We can then write an interpreter which is accepted by the typecheker
\begin{code}
let rec eval env = function
  | V VZ -> List.hd env
  | V (VS v) -> eval (List.tl env) (V v)
  | B b -> UB b
  | L e -> UA (fun x -> eval (x::env) e)
  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code}
and whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term:

@@

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, |List.hd| and
|List.tl| above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is when the first
operand is not a function.

Suppose our object language were typed too. Then terms like
|A (B true, B false)| would be ill-typed. Thus, if our evaluator
accepted only typed terms, the missing alternatives in the |A|
line would not occur. Alas, what is obvious to us is not obvious to
the meta-language. There is still a pattern match, and we still have
to attach and eliminate the UB and UA tags, even when there is no
longer a need for them. See \cite{WalidICFP02}, Section 1.4 for more details.

In short: It is not satisfactory
to use a Universal type because tagging hurts performance and
necessitates partial pattern matching in the metalanguage, even for a
well-typed object term.

\subsection{Other solutions}

But how should one design a type system to type
terms like |test1|? GADTs seem necessary. Similarly, how does one design a
typesystem that can keep track of the open/closed status of terms, so that we
can statically insure that only closed terms are passed to |eval| and that
the empty list checking in |List.hd| and |List.tl| can be eliminated.
The latter is an even more interesting problem requiring even
more advanced type systems 
(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
\jacques{I hope that is the right Davies paper}

This is also the main motivating problem (albeit in the context of
a first-order language) for the introduction of GADTs 
into Haskell \cite{PeytonJonesICFP2006}.  This was also a motivating
factor behind the work of Xi on GADTs \cite{XiGADTs}, and some of the
recent work on dependent types \cite{WalidICFP02,WalidConcoqtion}.

See the Related Work section (\ref{related}) for a discussion of the other
solutions to this problem.
\oleg{I feel we need to give some details of the other solutions here, 
to give the reader the feeling of
how advanced solutions were proposed and how rarely they are
implemented.}
\jacques{I have rearranged this section so that it should be clear how
to extend it to explain these other solutions}

\subsection{The ``Final'' Proposal}\label{ourapproach}

@@

Now, let us define our object language (embedded DSL) in a different
way. Instead of using algebraic datatypes to encode the terms of the
language in the meta-language (reminiscent to the \emph{initial algebra} 
approach), we will be using functions as term constructors (the \emph{final}
approach):

@@

\noindent That is all of the interpreter. Our sample term now reads
\begin{code}
  let testf1 = app (lam varZ) (b true)
\end{code}
which is almost the same as before, only written with lower-case
identifiers. To evaluate a term, we merely need to apply 
the term to the empty environment

@@

Furthermore, it should be clear that we do not use any
pattern-matching at all, and thus cannot have any pattern-matching
errors.  This is clear not just to us, but also to the compiler.
In other words in traditional approaches one can prove a meta-theorem
that all pattern-matching is total, but this is nevertheless not
apparent to the compiler, which thus still has to generate code for
those cases.  This seriously hurts efficiency~\cite{WalidICFP02}.

In the above code, the interpreter is wired in, directly built into the
functions |b|, |app|, etc.  In the remainder of this paper, 
we explain how to abstract the interpreter,
to make it possible to ``evaluate'' the same term in many different
ways: interpretation, compilation, compute the size (or depth), etc.

@@

   We have realized that if we change the encoding of the term --
from data constructors to functions, we can encode a tagless
interpreter of a typed language in a typed language with no GADTs and
a very simple type system. The interpreter uses no universal type, no
pattern matching and it is patently obvious to the compiler that the
only non-termination can come from interpreting the fixpoint
combinator |fix|. If the source
language is strongly normalizing, our interpreter is total.

@@

\noindent We have implemented the above in (Meta)OCaml and Haskell.
We view our contributions as:
\begin{enumerate}
\item we interpret/compile/partially-evaluate a typed source language
   from within a typed language.  The requirements of the host type
   system are \jacques{not so onerous?  simple?},
\item a simple, clean implementation,
\item a nicely comparable Haskell/MetaOCaml version,
\item a functorial approach which deals with uniformly with all
evaluators, from interpretation to partial evaluation to CBN CPS transform,
\item clear and easy extensibility to more features,
\item only well-typed and closed terms can be evaluated, and this is 
checked statically,
\item and an approach to self-interpretation which is compatible with the
  above.  Self-interpretation turned out to be considerably harder than
  we at first expected.
\end{enumerate}

We want to reiterate that it was previously thought that insuring
that types are well-typed \emph{or} closed required advanced type
extensions, and we show that this can be done in a simple subset
of both OCaml and Haskell.

\jacques{can we state fairly clearly what type system we need?  I mean,
Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
With our let-polymorphism trick for self-interpretation, what do we really
need?}

@@

The language we will embed is a simply-typed lambda-calculus
with fixpoint, integers, booleans and comparison, whose BNF grammar is

@@

This time we use higher-order abstract syntax \cite{DaleMiller}
rather than de Bruijn indices.  The
language is close to the language of \cite{XiPOPL03}, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Fibonacci, factorial, and power.

@@

\noindent |test1| now reads |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For example, we can encode the classical |power| function, as well
as the partial application |\x -> power x 7| as

\begin{code}
testpowfix () = 
  lam (\x -> fix (\self -> lam (\n ->
        if_ (leq n (int 0)) (int 1)
            (mul x (app self (add n (int (-1))))))))
testpowfix7 () = 
  lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code}
The dummy unit |()| argument above is to avoid the monomorphism
restriction and so that the type of |testpowfix| and |testpowfix7|
will remain polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signature.  We however prefer if the typechecker of the meta-language
(Haskell) infers the types of the object terms for us. We could also
avoid the |()| argument by switching off the monomophism restriction
with a compiler flag.

Note that the subject language is typed. If we make a mistake
(for example, replace |int 7| with |bool True| in |testpowfix7|, the
compiler will highlight that term and complain that it couldn't match
the expected |Int| against the inferred |Bool|. A type error
also results when entering |lam (\x -> app x x)|, with the compiler
complaining about an infinite type. The type-checker of the
meta-language will flag meaningless subject level terms --
before any of their evaluation. If we forget, for example, |app| in
some of the terms above, we get an error message with a precise
location. So the type-checker also flags syntactically invalid
subject-level terms.\oleg{Should we perhaps move this up, in the
  intro? To justify why typed-metalanguages and the typed DSL are
  good. Or keep the explanation here but refer to it from the Introduction.}

We have also embedded our DSL in (Meta)OCaml, replacing type classes with
functors, via a module type.  For this section, the following simple
signature will be sufficient:

@@

Some differences are in the additional type parameter |c| (the
environment classifier from \cite{WalidPOPL03}) to be used later when we get to
code generation (this is required by MetaOCaml as part of the type for the
code). MetaOCaml is a call-by-value language, and thus we have to 
$\eta$-expand
the fix-point combinator and encapsulate the branches of the |if_|
into thunks to prevent their evaluation.

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
     if_ (leq n (int 0)) (fun () -> int 1)
         (fun () -> mul x (app self 
                               (add n (int (-1))))))))
 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
The dummy unit argument in |testpowfix| is an artefact of MetaOCaml
(which we will be using to compile our tests in Sections below),
again related to monorphism. To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means the type
must be polymorhic. The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

This interpreter is patently tagless: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. We can run our
examples by instatiating the |EX| functor with |R|.
\noindent
\begin{code}
  module EXR = EX(R);;
\end{code}

\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define a\\
|newtype R a = R a| and |instance Symantics R|
with a straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during compilation) and pattern-matching
against |R| cannot ever fail and so is operationally an identity and
compiled that way.

Note that the interpreter uses no Universal type nor pattern
matching, and it is obvious to the compiler that non-termination can
only result from interpreting |fix|.

@@

\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
  let lam f        = (f 0) + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = (f 0) + 1
end;;
\end{code}

@@

We can also write a ``compiler'' from our lambda calculus directly
into OCaml code, using MetaOCaml's staging facilities. Namely,
future-stage expressions of the type |t| are represented at the
present stage as values of the type |('a,t) code| where |'a| is the
environment classifier \cite{WalidPOPL03,WalidESOP04}. Code values are created
by a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
evaluated at a future stage. The \emph{escape} |.~e|, which must occur
within a bracket, specifies that the expression |e| must be evaluated
at the current stage; its result, which must be a code value, is
spliced in into the code being built. The \emph{run} form |.! e|
evaluates the code value, which involves run-time compilation and
linking of the future-stage code. The bracket, escape and run are akin to
quasi-quotation, unquotation and |eval| of Lisp.

% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code
  let int (x:int)   = .<x>.
  let bool (b:bool) = .<b>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .< .~x <= .~y >.
  let if_ eb et ee = 
    .<if .~eb then .~(et () ) else .~(ee () )>.
  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f = 
     .<let rec self n = .~(f .<self>.) n in self>.
end;;
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler attempts no optimizations at all, it simply produces
residual code. For example, the evaluation of our |test1|
\begin{code}
  let module E = EX(C) in E.test1 ();;
\end{code}
gives a value with inferred type |('a, bool) C.repr| which looks
like |.<((fun x_6 -> x_6) (true))>.|. MetaOCaml can print the code
values, that is, the corresponding future-stage code.

It is readily obvious that this compiler does not suffer
any interpretive overhead whatsoever. Indeed, the
  code produced for a function is simply |fun x_6 -> x_6| and has no 
  embedded calls to the evaluator: cf. the evaluator of DSL functions
  in Section~\ref{tagproblem}, the L-line.
The resulting code obviously contains no tags and no pattern-matching.

We have also implemented this ``compiler'' in Haskell. Since Haskell
has no (convenient, typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with
constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
meaning. Unlike our staging language (which is based on higher-order
abstract syntax), our bytecode, to be realistic, uses integer-named
variables. We then define 
\begin{code}
  newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for the creation of fresh variable
names. We define the compiler by making |C| to be an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \textsf{ByteCode} \emph{datatype}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

@@

Surprisingly, it is also possible to write a partial evaluator!  The
partial evaluator does have tags: \emph{binding time} or \emph{phase}
tags, i.e. whether a value is static or dynamic.
Pattern-matching on this phase tag is always exhaustive; we still
have no universal type and no tags for object types.  The partial
evaluator uses the previously mentioned compiler to ``interpret'' dynamic
terms, and uses the interpreter to reduce static terms.  In this sense, 
the partial evaluator is a modular extension to both the compiler and
the interpreter.

\subsection{The trouble with (opaque) functions}
\label{S:PE-problem}
We want to combine the interpreter |R| and the compiler |C| to produce
an (online) partial evaluator.  We defer the discussion of the 
choice of online/offline partial evaluation until after we have presented our
solution in~\ref{S:PE-solution}.
The partial evaluator (PE) is also an interpreter: it
evaluates a term either to an immediate (or, `static' or present-stage)
value or to a `dynamic', future-stage value, or `code' (of the sort
produced by the compiler C). The interpreter attempts to perform all
possible computations (i.e., computations whose operarands are
present-stage values) as the present-stage. Our first attempt is
straightforward: we introduce
\begin{code}
data P1 t = S1 (R t) (C t) | E1 (C t)
\end{code}
as a `direct sum' of the interpreter |R| and the compiler |C|. The
datatype |P1| has two \emph{phase} (not type!) tags |S1| 
(for static) and |E1| for dynamic. The static value has both the
present-stage and future-stage alternatives. The dynamic value does
not have the present-stage component. We define
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (S1 _ dyn) = dyn
abstr1 (E1 dyn)   = dyn
\end{code}
to extract the future-stage component. We then attempt to define the
interpreter |P1| compositionally -- and it works for the first-order
fragment of our DSL language.

@@

\jacques{Comment on online/offline here}

@@

Boxes go Bananas by Weirich and Washburn.

@@

\bibliographystyle{plainnat}
\bibliography{tagless}
\end{document}
